Querying Integrated Scientific Observation and Measurement Data

### 2010-06-08 docs/algdraft/DataIntegration/method.tex
update the write-up for the algorithm method.

%Framework of query rewriting
Roughly speaking, the query rewriting consists of two steps:
\begin{itemize}
\item From the given query, search the database metadata and
  annotation information to find the relevant data tables and
  attributes that need to be used to answer the given query. 
\item Translate the given OM query to queries over the relevant data
  tables. 
\end{itemize}

@@

%Elaborate the first step
The first step is to map the OM concepts in the query to the real data
structure. 
With the annotation structures $OT$, $CT$, and $MT$, any given OM
concepts can be linked to a measurement type $mt_{id}$. 
Then, the annotation structure $Map$ can be used to bridge OM concepts and data sets. 
Utilizing $Map$, we can find the attributes $attr_i$ and the
annotation $A_{id}$).
E.g., when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $A_{id}$. 
In case that the given OM concept is a non-measurement concept, we can use
other annotation structures to figure out the measurementt types. 
E.g., if the OM concept is {\em characteristic} ``Species'', we can use
$MT$ to get the measurement type id $mt_{id}$. 
After we get $A_{id}$, we can quickly find the dataset id $d_{id}$
using $AD$ metadata table (Section \ref{sec:annotation}). 
Then, we get the relevant data tables and the needed attributes.

%Elaborate the second step
Let $D_{cand}$ be the set of relevant (or candidate) data tables and $d_{a_1} \cdots d_{a_m}$
are the related attributes for table $d\in D_{cand}$. 
The second step translates the original OM query to SQL queries over the data tables.

In particular, for each candidate data table $d$,  
the translation process works in the following rules.

\begin{itemize}
\item The SELECT clause always get the distinct $d_{id}$ and/or
  record id $r_{id}$ (when it is needed);
\item The FROM clause has table $d \in D_{cand}$;
\item The WHERE clause contains all the function $f(\cdot)$ in the OM
  query;
\item When there is distinct constraint on entity or observation, the
  GROUP BY clause contains of all the attributes corresponding to the
  {\em key measurement types} of the entity or observation. we detail
  all process of getting the key measurement types of an observation
  later. 
\item The aggregation requirement in $f(\cdot)$ is translated to
  HAVING clause. 
\end{itemize}

@@

\vspace{0.1in}
{\tt 
\begin{tabular}{l} 
SELECT DISTINCT $d_{id}$\\
FROM $d$\\
WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
$[$GROUP BY {\em distinct requirement}$]$\\
$[$HAVING {\em aggregation condition}$]$;
\end{tabular}
}
\vspace{0.1in}

@@

When the given query is a basic OM query, i.e., there is not context
in the query. 
Then, we can form a template query as the following;

>>>>

%Framework of query rewriting
Roughly speaking, the query rewriting consists of two steps:
\begin{itemize}
\item From the given query with entity, characteristic information, search the database metadata and
  annotation information to find the relevant data tables and
  attributes that need to be used to answer the given query.
\item Translate the given OM query to queries over the relevant data
  tables. To do this translation, we decompose an OM query to have several basic component operators.
  \begin{itemize}
  \item Basic OM query with one entity type. While the characteristic condition may be conjunctive or disjunctive, it may also has aggregation. E.g., $A(Cha1>25.0 \&\& Count(Cha)>3, Height>2.0)$
  \item Context OM chain refer to a set of queries having ``HasContext'' relationship. Obviously, these queries have different entity condition. E.g., $Ent1(...) \rightarrow Ent2(), Ent1() \rightarrow Ent3()$
  \item Disjunctive OM query with multiple entity types. $Ent1(...), Ent2(), Ent3(...)$.
  \end{itemize}

\end{itemize}

@@

%Elaborate the first step
The first step is to map the OM concepts in the query to the real data
structure.
With the annotation structures $OT$, $CT$, and $MT$, any given OM
concepts can be linked to a measurement type $mt_{id}$.
Then, the annotation structure $Map$ can be used to bridge OM concepts and data sets.
Utilizing $Map$, we can find the attributes $attr_i$ and the
annotation $A_{id}$).
E.g., when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $A_{id}$.
In case that the given OM concept is a non-measurement concept, we can use
other annotation structures to figure out the measurement types.
E.g., if the OM concept is {\em characteristic} ``Species'', we can use
$MT$ to get the measurement type id $mt_{id}$.
After we get $A_{id}$, we can quickly find the dataset id $d_{id}$
using $AD$ metadata table (Section \ref{sec:annotation}).
Then, we get the relevant data tables and the needed attributes.

%Elaborate the second step
Let $D_{cand}$ be the set of relevant (or candidate) data tables and $d_{a_1} \cdots d_{a_m}$
are the related attributes for table $d\in D_{cand}$.

The second step translates the original OM query to SQL queries over the data tables.
From now on, we assume that we know the attribute already for all the needed characteristics.

To translate the given query:
Figure \ref{alg:omq} shows the algorithm framework to deal with one OM query.
\from{HP}{More description comes later.}

\begin{figure}[htb]
%\normalsize
%\begin{minipage}{5in}
%\sf
\begin{scriptsize}
\begin{tabbing}
1166\= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= \kill
{\bf Algorithm Execute}(OMQ)\\
1).\> $Result$ := $\emptyset$;\\
2).\>//From the given query, get the DNF of context queries.\\
3).\> $contextChainDNF$ := ${\bf GroupBasicQuery}(OMQ)$;\\
4).\> \\
5).\> //Execute every DNF int the context chain set and union their results\\
6).\> {\bf for} each context chain $oneContextQuery$ in $contextChainDNF$\\
7).\>\> $result_{CNF}$ := $\emptyset$;\\
8).\>\>//Execute every basic query in this CNF query and intersect their results\\
9).\>\>{\bf for} each basic query $OM_{b}$ in $ContextQuery$\\
10).\>\>\>$result_{b}$ := ${\bf ExecuteBasicOM}(M_{b})$;\\
11).\>\>\>$result_{CNF}$ := $result_{CNF} \cap result_{b}$;\\
12).\>\>$Result$ := $Result\cup result_{CNF}$\\
13).\>Return $Result$;\\
\\
{\bf Algorithm GroupBasicQuery}($OMQ$);\\
1).\>$contextChainSet$ := $\emptyset$;\\
2).\\
3).\>//Put the basic queries in context into context chains\\
4).\>{\bf for} (each basic context $OM_{b}\rightarrow COM_{b}$ in $OMQ$)\\
5).\>\>$added$ := false;\\
6).\>\>{\bf for} (each set $oneContextChain \in contextChainDNF$)\\
7).\>\>\>{\bf if} (($OM_{b} \in oneContextChain$) or $OM_{b} \in oneContextChain$))\\
8).\>\>\>\> $oneContextChain$ := $oneContextChain \cup OM_{b} \cup COM_{b}$\\
9).\>\>\>\> $added$ :=true;\\
10).\>\>\>\> {\bf break;}\\
11).\>\>{\bf if}($added$==false)\\
12).\>\>\>Create a set $newContextChain$ with two elements $OM_{b}$ and $COM_{b}$;\\
13).\>\>\>Add $newContextChain$ to $contextChainSet$;\\
14).\\
15).\>//Put the basic queries without context into context chains\\
16).\>{\bf for} (each $OM_{b}$ not in context of $OMQ$)\\
17).\>\>Create a set $newContextChain$ with one element $OM_{b}$\\
18).\>\>Add $newContextChain$ to $contextChainSet$;\\
19).\>{\bf return} $contextChainSet$;
\end{tabbing}
\end{scriptsize}
%\end{minipage}
\caption{Algorithm to execute an OM query}
\label{alg:omq}
\end{figure}

In particular, for each candidate data table $d$,
the translation process works in the following rules.
For an OMBasic query with one entity type.
We need to do the

\begin{itemize}
\item The SELECT clause always get the distinct $d_{id}$ and the key attributes for this entity type;
\item The FROM clause has table $d \in D_{cand}$;
\item The WHERE clause contains all the non-aggregational conditions $f(\cdot)$ in the OM
  query;
\item When there is distinct constraint on entity or observation, the
  GROUP BY clause contains of all the key attributes corresponding to the of the entity. we detail
  all process of getting the key measurement types of an observation
  later.
\item The aggregation requirement in $f(\cdot)$ is translated to
  HAVING clause.
\end{itemize}

@@

For non-aggregate conditions,

\vspace{0.1in}
{\tt
\begin{tabular}{l}
SELECT DISTINCT $d_{id}, key~attribute~list$\\
FROM $d$\\
WHERE $non-aggregate~conditions$)
\end{tabular}
}
\vspace{0.1in}

For each aggregate condition,

\vspace{0.1in}
{\tt
\begin{tabular}{l}
SELECT DISTINCT $d_{id}, key~attribute~list$\\
FROM $d$\\
WHERE $key~attribute_values$ IN(\\
\hspace{0.6in}$non-aggregate~SQL$)\\
$[$GROUP BY {\em distinct requirement}$]$\\
$[$HAVING {\em aggregation condition}$]$;
\end{tabular}
}
\vspace{0.1in}

When there are more than one aggregation condition,
the above SQL is formed for each aggregation condition
and their results are intersected.

\from{HP}{Output: record id or key attribute values.}
If the output is record id, we further need to

\vspace{0.1in}
{\tt
\begin{tabular}{l}
SELECT $record\_id$\\
FROM $d$\\
WHERE ($key~measurement~value$ IN(\\
\hspace{0.5in}SELECT DISTINCT $key~attribute~list$\\
\hspace{0.5in}FROM ($results~from~aggregation functions$)
\end{tabular}
}
\vspace{0.1in}

@@

For a non-aggregate condition,

{\tt
\begin{tabular}{l}
SELECT DISTINCT did, record\_id, eid \\
\hspace{0.5in}characteristic, mvalue\\
FROM $mi, mt, oi, ei$\\
$[$WHERE    {\em table join condition}, \\
$\qquad\qquad$ AND {\em selection condition}$]$\\
INTERSECT $other~measurement~cond$\\
\end{tabular}
}

For an aggregate condition,

{\tt
\begin{tabular}{l}
SELECT DISTINCT did, record\_id, eid\\
FROM $oi$\\
WHERE (did, eid) IN (\\
\hspace{0.5in}SELECT did, eid\\
\hspace{0.5in}FROM $non-aggregate~sql$\\
\hspace{0.5in}WHERE $characteristic~condition$\\
\hspace{0.5in}GROUP BY did, eid\\
\hspace{0.5in}HAVING {\em aggregation condition};
\end{tabular}
}
When the given query is a basic OM query, i.e., there is not context
in the query.
Then, we can form a template query as the following;
